---
title: "Preprocessing"
author: "Alicia"
date: "2024-10-21"
output: html_document
output:
  pdf_document:
    toc: yes
    toc_depth: 2 
lang: es-ES
header-includes:
  - \usepackage{fvextra}\DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
  - "\\usepackage{geometry}"
  - "\\geometry{letterpaper, margin=1in}"
  - "\\usepackage{ragged2e}"
  - "\\justifying"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = FALSE, tidy = FALSE, 
               fig.width = 5, fig.height = 5,echo = TRUE, 
               message = FALSE, warning = FALSE, cache=FALSE, fig.align = "center")
Sys.setlocale("LC_TIME", "C")
```


# Step 1: Set working directory, create labels, and load relevant packages

```{r packages, echo = FALSE, message = FALSE}
options(warn = -1)
if(!(require(mlbench))) install.packages("mlbench")
if(!(require(caret))) install.packages("caret")
if(!(require(matrixStats))) install.packages("matrixStats")
if(!(require(hot.deck))) install.packages("hot.deck")
if(!(require(bnstruct))) install.packages("bnstruct")
if(!(require(plyr))) install.packages("plyr")
if(!(require(readxl))) install.packages("readxl")
if(!(require(knitr))) install.packages("knitr")
if(!(require(purrr))) install.packages("purrr")
if(!(require(ggplot2))) install.packages("ggplot2")
if(!require(car)) install.packages("car")
if(!require(factoextra)) install.packages("factoextra")
```

```{r libraries, echo = FALSE, message = FALSE}
library(caret)
library(matrixStats)
library(bnstruct)
library(plyr)
library(readxl)
library(dplyr)
library(purrr)
library(ggplot2)
library(car)
library(stats)
library(factoextra)
```

```{r}
setwd("C:/Users/alici/Desktop/TFM/R/Metabolomica")

today="041224"
cohort="PrePost_Menop"
ifilename="NMR_Metabolomic_Results_PinedaI001.xlsx"
clinic_data="datos_clinicos.xlsx"
imputed_filename=paste(today,cohort,"metabolomics","Imputed_risk_factors.txt",sep="_")
CM_filename=paste(today,cohort,"CorrelationMatrix_risk_factors.txt",sep="_")
cleaned_filename=paste(today,cohort,"Cleaned95_risk_factors.txt",sep="_")
cutoff=0.95


# Dataframe creation
sheet_names <- excel_sheets(ifilename)
data_list <- list()
for (sheet in sheet_names) {
  data <- read_excel(ifilename, sheet = sheet, skip = 6)
  data_list[[sheet]] <- data
}

metabolomic_data <- reduce(data_list, function(x, y) {
  inner_join(x, y, by = c("Sample ID BT", "Sample ID Client"))
})

head(metabolomic_data)

clinic_data <- read_excel(clinic_data)

metabolomic_data <- metabolomic_data %>%
  left_join(clinic_data, by = c("Sample ID BT", "Sample ID Client"))

head(metabolomic_data)
```


Next, we modify some variables:

```{r}
str(metabolomic_data)

# We change the year of birth to age
metabolomic_data$Edad <- 2024 - metabolomic_data$`Año nac.`
metabolomic_data$`Año nac.` <- NULL

# Convert categorical variables to numerical
metabolomic_data <- metabolomic_data %>%
  mutate(
    `Fumadora SI/NO` = case_when(
      `Fumadora SI/NO` %in% c("si", "sí", "SI") ~ 1,
      `Fumadora SI/NO` %in% c("no", "NO") ~ 2,
      TRUE ~ NA_real_
    ),
    `Presencia de diabetes SI/NO` = case_when(
      `Presencia de diabetes SI/NO` %in% c("si", "sí", "SI") ~ 1,
      `Presencia de diabetes SI/NO` %in% c("no", "NO") ~ 2,
      TRUE ~ NA_real_
    ),
    `Antecedentes de enfermedad cardiovascular SI/NO` = case_when(
      `Antecedentes de enfermedad cardiovascular SI/NO` %in% c("si", "sí", "SI") ~ 1,
      `Antecedentes de enfermedad cardiovascular SI/NO` %in% c("no", "NO") ~ 2,
      TRUE ~ NA_real_
    ),
    `Lugar de origen` = case_when(
      `Lugar de origen` == "España" ~ 1,
      TRUE ~ 2
    )
  )

# Transformed the variables into factors
metabolomic_data$`Fumadora SI/NO` <- factor(metabolomic_data$`Fumadora SI/NO`, levels = c(1, 2), labels = c("Sí", "No"))

metabolomic_data$`Presencia de diabetes SI/NO` <- factor(metabolomic_data$`Presencia de diabetes SI/NO`, levels = c(1, 2), labels = c("Sí", "No"))

metabolomic_data$`Antecedentes de enfermedad cardiovascular SI/NO` <- factor(metabolomic_data$`Antecedentes de enfermedad cardiovascular SI/NO`, levels = c(1, 2), labels = c("Sí", "No"))

metabolomic_data$`Lugar de origen` <- factor(metabolomic_data$`Lugar de origen`, levels = c(1, 2), labels = c("España", "Otro"))

head(metabolomic_data["Fumadora SI/NO"])

metabolomic_data$`Grupo_factores` <- factor(metabolomic_data$`Grupo_factores`, levels = c(1, 2, 3), labels = c("1", "2", "3"))

# Check the distribution of the factors
table(metabolomic_data["Fumadora SI/NO"])
table(metabolomic_data["Presencia de diabetes SI/NO"])
table(metabolomic_data["Antecedentes de enfermedad cardiovascular SI/NO"])
table(metabolomic_data["Lugar de origen"])
table(metabolomic_data["Grupo_factores"])

# Rename variables
metabolomic_data <- metabolomic_data %>%
  dplyr::rename(Grupo = "Grupo_factores",
         VLDL_C = "VLDL-C",
         IDL_C = "IDL-C",
         LDL_C = "LDL-C",
         HDL_C = "HDL-C",
         VLDL_TG = "VLDL-TG",
         IDL_TG = "IDL-TG",
         LDL_TG = "LDL-TG",
         HDL_TG = "HDL-TG",
         Total_VLDL_P = "VLDL-P (nmol/L)",
         L_VLDL_P = "Large VLDL-P (nmol/L)",
         M_VLDL_P = "Medium VLDL-P (nmol/L)",
         S_VLDL_P = "Small VLDL-P (nmol/L)",
         Total_LDL_P = "LDL-P (nmol/L)",
         L_LDL_P = "Large LDL-P (nmol/L)",
         M_LDL_P = "Medium LDL-P (nmol/L)",
         S_LDL_P = "Small LDL-P (nmol/L)",
         Total_HDL_P = "HDL-P (μmol/L)",
         L_HDL_P = "Large HDL-P (μmol/L)",
         M_HDL_P = "Medium HDL-P (μmol/L)",
         S_HDL_P = "Small HDL-P (μmol/L)",
         VLDL_Z = "VLDL-Z (nm)",
         LDL_Z = "LDL-Z (nm)",
         HDL_Z = "HDL-Z (nm)",
         Glyc_A = "Glyc-A (μmol/L)",
         Glyc_B = "Glyc-B (μmol/L)",
         Glyc_F = "Glyc-F (μmol/L)",
         HW_Glyc_B = "H/W Glyc-B",
         HW_Glyc_A = "H/W Glyc-A",
         "3_Hydroxybutyrate" = "3-Hydroxybutyrate",
         Glucose = "Glucose(mmol/L)",
         w6_w7 = "w6+w7",
         ARA_EPA = "ARA+EPA",
         Diabetes = "Presencia de diabetes SI/NO",
         Antecedentes = "Antecedentes de enfermedad cardiovascular SI/NO",
         Tension_sistolica = "Tensión sistólica (mmHg)",
         Origen = "Lugar de origen",
         Fumadora = "Fumadora SI/NO",
         Total_PUFA = "PUFA",
         Total_SFA = "SFA"
         )

colnames(metabolomic_data)
str(metabolomic_data)
summary(metabolomic_data)
```

Create 5 new variables: Total choleterol (Total C), ratio Cholesterol/HDL (atherogenic index), Total Triglycerides, TG/PG y HDL/LDL.

```{r}
# Calculate total cholesterol
metabolomic_data$"Total_C" <- metabolomic_data$"VLDL_C" + metabolomic_data$"IDL_C" + metabolomic_data$"LDL_C" + metabolomic_data$"HDL_C"

# Calculate the Total C / HDL-C ratio
metabolomic_data$"Total_C_HDL_C" <- metabolomic_data$"Total_C" / metabolomic_data$"HDL_C"

# Calculate total triglycerides
metabolomic_data$"Total_Triglycerides" <- metabolomic_data$"VLDL_TG" + metabolomic_data$"IDL_TG" + metabolomic_data$"LDL_TG" + metabolomic_data$"HDL_TG"

# Calculate the TG / PG ratio
metabolomic_data$"TG_PG" <- metabolomic_data$"Total_Triglycerides" / (metabolomic_data$"PL" + metabolomic_data$"LPC")

# Calculate the HDL / LDL ratio
metabolomic_data$"HDL_LDL" <- metabolomic_data$"HDL_C" / metabolomic_data$"LDL_C"

metabolomic_data[,72:77]
```


Analyse missing values:

```{r}
metabolomic_data[metabolomic_data == "NA"] <- NA
metabolomic_data[metabolomic_data == "<LOD"] <- NA
anyNA(metabolomic_data)

missing_values <- sapply(metabolomic_data, function(x) sum(is.na(x)))
print(missing_values)
nrow(metabolomic_data)
```

Removal of variables with too many missing values:

```{r}
columns_to_remove <- c("3_Hydroxybutyrate")

metabolomic_data <- metabolomic_data %>% 
  select(-all_of(columns_to_remove))

head(metabolomic_data)
str(metabolomic_data)
```

```{r}
columns_to_impute <- metabolomic_data %>%
  select(Acetate, Glycerol, Threonine)
matrix_metabolomic_data=as.matrix(columns_to_impute) # For knn.impute only, otherwise leave it as a data.frame
```


## Step 2: Impute missing values

### (Optional jump in at the end of step 2 if you have no missing data)

```{r}
imputed=knn.impute(matrix_metabolomic_data)

write.table(imputed,imputed_filename,row.names=T,col.names=T,quote=F,sep="\t")

write.table(imputed,"imputed_metabolomic_data.csv",row.names=T,col.names=T,quote=F,sep=",")
```

####
####  Already have data complete without missing data? Read in here! (remove the hash)
####

```{r}
i_filename="041224_PrePost_Menop_metabolomics_Imputed_risk_factors.txt"
imputed_metabolomic_data=read.table(i_filename,as.is=T,row.names=1,h=T,check.names=F,sep="\t")
anyNA(imputed_metabolomic_data)
```

Compare the statistics before and after the imputation:

```{r}
compare_statistics <- function(data1, var1, data2, var2) {
  if (!(var1 %in% colnames(data1))) {
    stop(paste("La variable", var1, "no está en metabolomic_data"))
  }
  if (!(var2 %in% colnames(data2))) {
    stop(paste("La variable", var2, "no está en metabolomic_data"))
  }

  data1[[var1]] <- as.numeric(data1[[var1]]) 
  data2[[var2]] <- as.numeric(data2[[var2]])
  
  stats_var1 <- summary(data1[[var1]], na.rm = TRUE)
  stats_var1 <- data.frame(Statistic = names(stats_var1), Value = as.numeric(stats_var1))
  
  stats_var2 <- summary(data2[[var2]], na.rm = TRUE)
  stats_var2 <- data.frame(Statistic = names(stats_var2), Value = as.numeric(stats_var2))
  
  comparison <- merge(stats_var1, stats_var2, by = "Statistic", suffixes = c("_original", "_imputed"))
  
  return(comparison)
}

# Statistics after imputation
statistics_acetate <- compare_statistics(metabolomic_data, "Acetate", imputed_metabolomic_data, "Acetate")
print(statistics_acetate)

statistics_glycerol <- compare_statistics(metabolomic_data, "Glycerol", imputed_metabolomic_data, "Glycerol")
print(statistics_glycerol)

statistics_threonine <- compare_statistics(metabolomic_data, "Threonine", imputed_metabolomic_data, "Threonine")
print(statistics_threonine)
```


```{r}
imputed_columns_df <- as.data.frame(imputed_metabolomic_data)

metabolomic_data_imputed <- metabolomic_data %>%
  select(-Acetate, -Glycerol, -Threonine) %>%
  bind_cols(imputed_columns_df)

head(metabolomic_data_imputed)
anyNA(metabolomic_data_imputed)
```




## Step 3: Calculate correlation matrix

```{r}
toCor=metabolomic_data_imputed[, -c(1, 2, which(colnames(metabolomic_data_imputed) %in% c("Grupo", "Antecedentes", "Diabetes", "Fumadora", "Origen", "Nodo", "Tension_sistolica", "Total_C", "Total_Triglycerides")))]
# Get rid of columns with clinical features, phenotype, or sampleID prior to doing the correlation matrix
str(toCor)
cov=metabolomic_data_imputed[, c(1, 2, which(colnames(metabolomic_data_imputed) %in% c("Grupo", "Antecedentes", "Diabetes", "Fumadora", "Origen", "Nodo", "Tension_sistolica","Total_C", "Total_Triglycerides")))]

correlationMatrix <- cor(toCor)
diag(correlationMatrix)<-0

write.table(correlationMatrix,CM_filename,row.names=T,col.names=T,quote=F,sep="\t")
```


## Step 4: Homology reduction

```{r}
metabolomic_data_imputed_cleaned <- metabolomic_data_imputed %>%
  select(-Nodo, -"Sample ID BT", -"Sample ID Client", -"Total_C", -"Total_Triglycerides", -"Fumadora", -"Tension_sistolica", -c(, "Antecedentes", "Diabetes", "Origen"))

correlationMatrix <- abs(correlationMatrix)

hc <- caret::findCorrelation(correlationMatrix, cutoff=cutoff)
##:: can be used to call a function from a package. In this case I am telling R to use findCorrelation from the caret package

columns_to_remove <- colnames(toCor)[hc]
print(columns_to_remove)

str(metabolomic_data_imputed_cleaned)

metabolomic_data_cleaned <- metabolomic_data_imputed_cleaned %>%
  select(-all_of(columns_to_remove))

str(metabolomic_data_cleaned)
```


## Step 6: Write table

```{r}
write.table(metabolomic_data_cleaned,cleaned_filename,row.names=T,col.names=T,quote=F,sep="\t")

columnas_iniciales <- colnames(metabolomic_data_imputed_cleaned)
columnas_finales <- colnames(metabolomic_data_cleaned)
ncol(metabolomic_data_imputed_cleaned)
ncol(metabolomic_data_cleaned)

columnas_eliminadas <- setdiff(columnas_iniciales, columnas_finales)

print("Columnas eliminadas:")
print(columnas_eliminadas)

str(metabolomic_data_cleaned)
```




# Analysis of outliers

Outliers per variable: 

```{r}
if (!dir.exists("Outliers")) {
  dir.create("Outliers")
}

counter <- 1

for (column in names(cleaned_ids)) {
  if (column != "Grupo" && column != "Sample ID Client")
    
    cleaned_data <- cleaned_ids[!is.na(cleaned_ids[[column]]), ]
    cleaned_data$Grupo <- as.factor(cleaned_data$Grupo)
    
    Q1 <- tapply(cleaned_data[[column]], cleaned_data$Grupo, function(x) quantile(x, 0.25))
    Q3 <- tapply(cleaned_data[[column]], cleaned_data$Grupo, function(x) quantile(x, 0.75))
    IQR <- Q3 - Q1
    lower_bound <- Q1 - 1.5 * IQR
    upper_bound <- Q3 + 1.5 * IQR
    
    outliers_data <- cleaned_data[
      cleaned_data[[column]] < unlist(lower_bound)[as.factor(cleaned_data$Grupo)] | 
      cleaned_data[[column]] > unlist(upper_bound)[as.factor(cleaned_data$Grupo)], 
    ]
    
    outliers_data$Sample_ID <- cleaned_data$`Sample ID Client`[match(outliers_data$`Sample ID Client`, cleaned_data$`Sample ID Client`)]
    
    p <- ggplot(cleaned_data, aes(x = Grupo, y = .data[[column]], fill = Grupo)) +
      geom_boxplot(outlier.shape = NA, alpha = 0.5) +  
      scale_fill_manual(values = c("skyblue", "lightgreen", "lightpink")) + 
      geom_point(data = outliers_data, aes(x = Grupo, y = .data[[column]]), color = "red", size = 3) +
      geom_text(data = outliers_data, aes(x = Grupo, y = .data[[column]], label = Sample_ID), 
                color = "red", hjust = 0.5, vjust = -0.5, size = 2) +  
      labs(title = paste("Boxplot for", column), 
           x = "Grupo", 
           y = column) +
      theme_minimal() +
      theme(legend.position = "none",  # Eliminar la leyenda
            axis.title.x = element_text(size = 14),
            axis.title.y = element_text(size = 14),
            plot.title = element_text(size = 16, face = "bold"))
    
    ggsave(filename = paste0("Outliers/boxplot_", counter, ".png"), plot = p, width = 8, height = 6, dpi = 300, bg = "white")
    
    counter <- counter + 1
  }
}
```


Outliers per sample:

```{r}
outlier_per_sample_list <- list()

for (column in names(cleaned_ids)) {
  if (column != "Grupo" && column != "Sample ID Client") {
    cleaned_data <- cleaned_ids[!is.na(cleaned_ids[[column]]), ]  
    cleaned_data$Grupo <- as.factor(cleaned_data$Grupo)
    Q1 <- tapply(cleaned_data[[column]], cleaned_data$Grupo, function(x) quantile(x, 0.25))
    Q3 <- tapply(cleaned_data[[column]], cleaned_data$Grupo, function(x) quantile(x, 0.75))
    IQR <- Q3 - Q1
    lower_bound <- Q1 - 1.5 * IQR
    upper_bound <- Q3 + 1.5 * IQR
    
    # Identifiy outliers
    outliers_data <- cleaned_data[
      cleaned_data[[column]] < unlist(lower_bound)[as.factor(cleaned_data$Grupo)] | 
      cleaned_data[[column]] > unlist(upper_bound)[as.factor(cleaned_data$Grupo)], 
    ]
    
    if (nrow(outliers_data) > 0) {
      outliers_data$Sample_ID <- cleaned_data$`Sample ID Client`[match(outliers_data$`Sample ID Client`, cleaned_data$`Sample ID Client`)]
      for (i in 1:nrow(outliers_data)) {
        sample_id <- outliers_data$Sample_ID[i]
        if (!(sample_id %in% names(outlier_per_sample_list))) {
          outlier_per_sample_list[[sample_id]] <- data.frame("Sample ID Client" = sample_id, stringsAsFactors = FALSE)
        }
        outlier_per_sample_list[[sample_id]][, column] <- outliers_data[[column]][i]
      }
    }
  }
}

for (sample_id in names(outlier_per_sample_list)) {
  print(outlier_per_sample_list[[sample_id]])
}

```


Tables with the range and mean of each variable:

```{r}
summary_tables <- list()

for (column in names(cleaned_ids)) {
  if (column != "Grupo" && column != "Sample ID Client") {  
    media <- mean(cleaned_ids[[column]], na.rm = TRUE)
    min_value <- min(cleaned_ids[[column]], na.rm = TRUE)
    max_value <- max(cleaned_ids[[column]], na.rm = TRUE)
    
    variable_summary <- data.frame("Variable" = column, 
                                   "Media" = media, 
                                   "Min" = min_value, 
                                   "Max" = max_value)
    summary_tables[[column]] <- variable_summary
  }
}

for (table_name in names(summary_tables)) {
  cat("\nSummary table for the variable", table_name, ":\n")
  print(summary_tables[[table_name]])
}


```



Analyse batch effect: 

```{r}
nodes_data <- read_excel("C:/Users/alici/Desktop/TFM/R/Metabolomica/datos_clinicos.xlsx")

nodes_df <- cleaned_ids %>%
  left_join(nodes_data %>% select(`Sample ID Client`, Nodo), by = "Sample ID Client")

nodes_df$Nodo <- as.factor(nodes_df$Nodo)

pca_result <- prcomp(nodes_df %>% select(where(is.numeric)), center = TRUE, scale. = TRUE)

summary(pca_result)

fviz_eig(pca_result)

p <- fviz_pca_ind(pca_result,
                  geom.ind = "point",
                  col.ind = nodes_df$Nodo,
                  palette = c("#00AFBB", "#E7B800", "#FC4E07", "#7CFC00", "#FF1493"),
                  addEllipses = TRUE,
                  ellipse.type = "convex",
                  legend.title = "Node",
                  pointsize = 2) + 
     labs(title = "PCA by Node") 

p <- p + scale_shape_manual(values = rep(16, length(unique(nodes_df$Nodo))))

ggsave("pca.svg", plot= p,width = 6, height = 4, dpi = 300)
```