---
title: "XGBoost TFM"
author: "Alicia"
date: "2024-12-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r directory}
setwd("C:/Users/alici/Desktop/TFM/R/Metabolomica")
```

```{r packages}
install.packages("xgboost")
```

```{r libraries}
library(caret)
library(ROCR)
library(MLeval)
library(readxl)
library(xgboost)
```

```{r}
g1="Pre_M_LR"
g2="Post_M_LR" 
g3="Post_M_HR"
compar1=paste(g1,"vs",g2,sep="")
compar2=paste(g2,"vs",g3,sep="")
today="161224"
label1=paste(today,"_",compar1,"_XGBoost",sep="")
label2=paste(today,"_",compar2,"_XGBoost",sep="")
classes1 <- c("Pre_M_LR", "Post_M_LR")
classes2 <- c("Post_M_LR", "Post_M_HR")
cmOut1=paste(label1,"_ConfusionMatrix.txt",sep="")
cmOut2=paste(label2,"_ConfusionMatrix.txt",sep="")

data=read.table("C:/Users/alici/Desktop/TFM/R/Metabolomica/041224_PrePost_Menop_Cleaned95_risk_factors.txt",as.is=T,h=T,check.names=T,sep="\t",na = c("NA", "NDEF", "TAG","NaN","ND"))
str(data)
```


# Prepare data for the analysis of Post-M-LR vs Pre-M-LR

```{r}
data$Grupo <- factor(data$Grupo, levels= c("1", "2", "3"), labels= c("Pre_M_LR", "Post_M_LR", "Post_M_HR"))
data_g1_g2 <- data[data$Grupo %in% classes1, ]
data_g1_g2 <- droplevels(data[data$Grupo %in% classes1, ])
table(data_g1_g2$Grupo)
levels(data_g1_g2$Grupo)
#str(data_g1_g2)

data_g1_g2$Edad <- as.numeric(data_g1_g2$Edad)


y=as.matrix(data_g1_g2$Grupo)
y=as.factor(y)


clinical <- data_g1_g2[, !colnames(data_g1_g2) %in% c("Grupo", "Edad")]
#str(clinical)

prepro=preProcess(clinical,method=c("center","scale")) 

x=predict(prepro,clinical)

#str(x)
anyNA(x)

x=data.matrix((x))
```

# Prepare data for the analysis of Post-M-HR vs Post-M-LR

```{r}
data_g2_g3 <- data[data$Grupo %in% classes2, ]
data_g2_g3 <- droplevels(data[data$Grupo %in% classes2, ])
table(data_g2_g3$Grupo)
levels(data_g2_g3$Grupo)
#str(data_g2_g3)

data_g2_g3$Edad <- as.numeric(data_g2_g3$Edad)

y2=as.matrix(data_g2_g3$Grupo)
y2=as.factor(y2)


clinical2 <- data_g2_g3[, !colnames(data_g2_g3) %in% c("Grupo")]
#str(clinical2)

prepro2=preProcess(clinical2,method=c("center","scale")) 

x2=predict(prepro2,clinical2)
str(x2)
anyNA(x2)

x2=data.matrix((x2))
```


# Function for obtaining the AUC-ROC curve with TPR and FPR

```{r}
plot_roc_auc <- function(optimized_model, main_title = "ROC Curve", positive_class = NULL) {
  # Load necessary libraries
  if (!requireNamespace("pROC", quietly = TRUE)) install.packages("pROC")
  if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
  library(pROC)
  library(ggplot2)
  
  # Extract predictions
  pred <- optimized_model$pred
  obs <- as.factor(pred$obs)
  
  # Check if the positive class exists in the levels
  if (!positive_class %in% levels(obs)) {
    stop("The specified positive class is not present in the data.")
  }
  
  # Reorder levels of the observed variable to ensure the positive class is the one of interest
  obs <- factor(obs, levels = c(positive_class, setdiff(levels(obs), positive_class)))
  
  # Extract predicted probabilities for the positive class
  predicted_probabilities <- as.numeric(pred[[positive_class]])
  
  # Calculate ROC curve
  roc_curve <- roc(obs, predicted_probabilities, levels = rev(levels(obs)))
  auc_value <- auc(roc_curve)
  
  # Extract TPR and FPR
  roc_data <- data.frame(
    TPR = roc_curve$sensitivities,         # True Positive Rate
    FPR = 1 - roc_curve$specificities      # False Positive Rate
  )
  
  # Plot TPR and FPR
  plot(roc_data$FPR, roc_data$TPR, type = "l", col = "red", lwd = 2, 
       xlab = "False Positive Rate", ylab = "True Positive Rate", 
       main = main_title)
  abline(a = 0, b = 1, col = "gray", lty = 2)  # Random diagonal
  
  # Add legend
  legend("bottomright", 
         legend = paste(positive_class, "\nAUC-ROC =", round(auc_value, 3)), 
         col = "red", 
         lwd = 2, 
         bty = "n", 
         text.col = "black")
  
  # Return AUC in case it is needed
  return(auc_value)
}
```



# Apply XGBoost for Post-M-LR vs Pre-M-LR: 

```{r}
set.seed(10)

Train_control <- trainControl(method="cv",
                              number=5,
                              summaryFunction = twoClassSummary,
                              savePredictions = "final",
                              classProbs = TRUE,
                              sampling = "smote")


xgbGrid <- expand.grid(nrounds = c(100,200),
                       max_depth = c(10, 15, 20, 25),
                       colsample_bytree = seq(0.5, 0.9, length.out = 5),
                       eta = 0.1,
                       gamma=0,
                       min_child_weight = 1,
                       subsample = 1
)


xgb_model_g1g2 = train(
  x, y,  
  trControl = Train_control,
  tuneGrid = xgbGrid,
  method = "xgbTree",
  strata=data_g1_g2$Grupo
)

xgb_model_g1g2$results
xgb_model_g1g2$bestTune

conf_matrix <- confusionMatrix(
  data = xgb_model_g1g2$pred$pred,  
  reference = xgb_model_g1g2$pred$obs,  
  positive = "Post_M_LR"
)

print(conf_matrix)
```


# AUC-ROC curve and confusion matrix 

```{r}
# AUC-ROC curve
auc_rf <- plot_roc_auc(
  xgb_model_g1g2, 
  main_title = "XGB", 
  positive_class = "Post_M_LR"
)

svg("roc_auc_XGB_g1g2.svg", width = 4, height = 4)

plot_roc_auc(
  xgb_model_g1g2, 
  main_title = "XGB", 
  positive_class = "Post_M_LR"
)

dev.off() 

# Confusion matrix
confusion <- table(Predicted = xgb_model_g1g2$pred$pred, Actual = xgb_model_g1g2$pred$obs)

conf_matrix_df <- as.data.frame(confusion)

print(conf_matrix_df)

colnames(conf_matrix_df) <- c("Prediction", "Reference", "Freq")

conf_matrix_df$Prediction <- factor(conf_matrix_df$Prediction, levels = c("Pre_M_LR", "Post_M_LR"))
conf_matrix_df$Reference <- factor(conf_matrix_df$Reference, levels = c("Post_M_LR", "Pre_M_LR"))

conf_matrix_df$Class <- ifelse(conf_matrix_df$Prediction == conf_matrix_df$Reference, "True", "False")

confusion_plot <- ggplot(conf_matrix_df, aes(x = Reference, y = Prediction, fill = Class)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 7) +
  scale_fill_manual(
    values = c("True" = "#A8E6A2", "False" = "#F4B6B2"),  
    name = "Classification",                              
    labels = c("False", "True")
  ) +
  scale_x_discrete(position = "top") +  
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 11)
  )

print(confusion_plot)

ggsave("conf_matrix_XGB_g1g2.svg", plot = confusion_plot, width = 6, height = 2, dpi = 300)
```


# Calculate SHAP values for the model 

```{r}
x <- as.data.frame(x)

# Create the predictor object for the fitted model
predictor <- Predictor$new(
  model = xgb_model_g1g2,       
  data = x,                       
  y = y                           
)

shap_values_all <- list()

# Iterate over all observations
for (i in 1:nrow(x)) {
  cat("Calculating SHAP for sample", i, "of", nrow(x), "\n")
  
  x_interest_single <- x[i, , drop = FALSE]
  
  shapley <- Shapley$new(predictor, x.interest = x_interest_single)
  
  shap_values <- shapley$results
  shap_values$sample_id <- i
  
  shap_values_all[[i]] <- shap_values
}

# Combine all results into a single dataframe
shap_values_df <- bind_rows(shap_values_all)

# Calculate the average importance of each variable
shap_summary <- shap_values_df %>%
  group_by(feature) %>%
  summarize(mean_phi = mean(abs(phi)), .groups = "drop") %>%
  arrange(desc(mean_phi))

# Select the top 10 most important variables
top_10 <- shap_summary[1:10, ]

# Visualize the most important variables globally
shap_XGB_g1g2 <- ggplot(top_10, aes(x = reorder(feature, mean_phi), y = mean_phi)) +
  geom_bar(stat = "identity", fill = "#78B056") +
  coord_flip() +
  labs(title = "Average importance of variables",
       x = "Metabolites",
       y = "Average SHAP value (|phi|)") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.line = element_line(color = "black"),  
        axis.ticks = element_line(color = "black"))  

# Save the image
ggsave("shap_importance_plot_XGB_g1g2.svg", plot = shap_XGB_g1g2, width = 8, height = 4, dpi = 300)
```



# Violin plot of the 10 top performing metabolites

```{r}
# List of variables to plot
variables <- c("SM", "Glutamine", "Creatinine", "S_LDL_P", 
               "Glucose", "FC", "EC", "VLDL_Z", "w3", "LDL_Z")

# List of units for each variable
units <- c("mmol/L", "μmol/L", "μmol/L", "nmol/L", 
           "mmol/L", "mmol/L", "mmol/L", "nm", 
           "mmol/L", "nm")

plot_list <- list()

data_g1_g2$Grupo <- factor(data_g1_g2$Grupo, levels = c("Pre_M_LR", "Post_M_LR"))

# Loop to generate violin plots
for (i in 1:length(variables)) {
  
  # Calculate p-value with t-test
  p_value <- t.test(data_g1_g2[[variables[i]]] ~ data_g1_g2$Grupo)$p.value
  if (p_value < 0.001) {
    p_label <- "p < 0.001"
  } else {
    p_label <- paste0("p = ", round(p_value, 3))
  }
  
  # Create the plot with the corresponding unit
  p <- ggplot(data_g1_g2, aes_string(x = "Grupo", y = variables[i], fill = "Grupo")) +
  geom_violin(trim = FALSE, width = 1.2, color = "black", alpha = 0.6) +
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.7, color = "black") +
  stat_summary(fun = "median", geom = "point", size = 2, color = "black") + 
  labs(x = NULL, y = units[i], title = variables[i]) +
  annotate("text", x = 1.5, y = Inf, label = p_label, vjust = 2, size = 8, fontface = "plain") +
  theme_minimal(base_size = 14) +  
  scale_fill_manual(values = c("#A7A7A7", "#F78D80"), guide = "none") + 
  scale_x_discrete(labels = levels(data_g1_g2$Grupo)) +  
  theme(
    legend.position = "none", 
    axis.title = element_text(size = 20),  
    axis.text.x = element_text(size = 20),  
    axis.ticks.x = element_blank(),  
    plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),  
    plot.margin = margin(5, 5, 10, 5) 
  )
  
  plot_list[[variables[i]]] <- p
}

combined_plot <- ggarrange(plotlist = plot_list, ncol = 5, nrow = 2)

print(combined_plot)

# Save the combined image
ggsave("Violin_Plots_XGB_g1g2.svg", combined_plot, width = 20, height = 10, bg = "white")
```





# Apply XGBoost for Post-M-HR vs Post-M-LR:

```{r}
set.seed(10)

Train_control <- trainControl(method="cv",
                              number=5,
                              summaryFunction = twoClassSummary,
                              savePredictions = "final",
                              classProbs = TRUE,
                              sampling = "smote")


xgbGrid <- expand.grid(nrounds = c(100,200),
                       max_depth = c(10, 15, 20, 25),
                       colsample_bytree = seq(0.5, 0.9, length.out = 5),
                       eta = 0.1,
                       gamma=0,
                       min_child_weight = 1,
                       subsample = 1
)


xgb_model_g2g3 = train(
  x2, y2,  
  trControl = Train_control,
  tuneGrid = xgbGrid,
  method = "xgbTree",
  strata=data_g2_g3$Grupo
)

xgb_model_g2g3$results
xgb_model_g2g3$bestTune

conf_matrix <- confusionMatrix(
  data = xgb_model_g2g3$pred$pred,
  reference = xgb_model_g2g3$pred$obs,  
  positive = "Post_M_HR"
)

print(conf_matrix)
```

# AUC-ROC curve and confusion matrix 

```{r}
# AUC-ROC curve 
auc_rf <- plot_roc_auc(
  xgb_model_g2g3, 
  main_title = "XGB", 
  positive_class = "Post_M_HR"
)

svg("roc_auc_XGB_g2g3.svg", width = 4, height = 4)

plot_roc_auc(
  xgb_model_g2g3, 
  main_title = "XGB", 
  positive_class = "Post_M_HR"
)

dev.off() 

# Confusion matrix
confusion <- table(Predicted = xgb_model_g2g3$pred$pred, Actual = xgb_model_g2g3$pred$obs)

conf_matrix_df <- as.data.frame(confusion)

print(conf_matrix_df)

colnames(conf_matrix_df) <- c("Prediction", "Reference", "Freq")

conf_matrix_df$Prediction <- factor(conf_matrix_df$Prediction, levels = c("Post_M_LR", "Post_M_HR"))
conf_matrix_df$Reference <- factor(conf_matrix_df$Reference, levels = c("Post_M_HR", "Post_M_LR"))

conf_matrix_df$Class <- ifelse(conf_matrix_df$Prediction == conf_matrix_df$Reference, "True", "False")

confusion_plot <- ggplot(conf_matrix_df, aes(x = Reference, y = Prediction, fill = Class)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 7) +
  scale_fill_manual(
    values = c("True" = "#A8E6A2", "False" = "#F4B6B2"),
    name = "Classification",   
    labels = c("False", "True")
  ) +
  scale_x_discrete(position = "top") +
  labs(
    title = "Matriz de Confusión",
    x = "Actual",
    y = "Predicted"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 11)
  )

print(confusion_plot)

ggsave("conf_matrix_XGB_g2g3.svg", plot = confusion_plot, width = 6, height = 2, dpi = 300)
```

# Calculate SHAP values for the model 

```{r}
x2 <- as.data.frame(x2)

# Create the predictor object for the fitted model
predictor <- Predictor$new(
  model = xgb_model_g2g3,       
  data = x2,        
  y = y2                      
)

shap_values_all <- list()

# Iterate over all observations
for (i in 1:nrow(x2)) {
  cat("Calculando SHAP para muestra", i, "de", nrow(x2), "/n")
  
  x_interest_single <- x2[i, , drop = FALSE]
  
  shapley <- Shapley$new(predictor, x.interest = x_interest_single)
  
  shap_values <- shapley$results
  shap_values$sample_id <- i
  
  shap_values_all[[i]] <- shap_values
}

# Combine all results into a single dataframe
shap_values_df <- bind_rows(shap_values_all)

# Calculate the average importance of each variable
shap_summary <- shap_values_df %>%
  group_by(feature) %>%
  summarize(mean_phi = mean(abs(phi)), .groups = "drop") %>%
  arrange(desc(mean_phi))

# Select the top 10 most important variables
top_10 <- shap_summary[1:10, ]

# Visualize the most important variables globally
shap_XGB_g2g3 <- ggplot(top_10, aes(x = reorder(feature, mean_phi), y = mean_phi)) +
  geom_bar(stat = "identity", fill = "#78B056") +
  coord_flip() +
  labs(title = "Average importance of variables (SHAP)",
       x = "Features (Metabolites)",
       y = "Average SHAP value (|phi|)") +
  theme_minimal() +
  theme(panel.grid = element_blank())  

# Save the image
ggsave("shap_importance_plot_XGB_g2g3.png", plot = shap_XGB_g2g3, width = 8, height = 4, dpi = 300)

```


# Violin plot of the 10 top performing metabolites

```{r}
# List of variables to plot
variables <- c("EC", "FC", "LDL_TG", "Valine", 
               "IDL_C", "Glycine", "HDL_C", "Total_C_HDL_C", "VLDL_Z", "Glyc_A")

# List of units for each variable
units <- c("mmol/L", "mmol/L", "mg/dL", "μmol/L", 
           "mg/dL", "μmol/L", "mg/dL", "Ratio", 
           "nm", "μmol/L")

plot_list <- list()

# Loop to generate violin plots
for (i in 1:length(variables)) {
  
  # Calculate p-value with t-test
  p_value <- t.test(data_g2_g3[[variables[i]]] ~ data_g2_g3$Grupo)$p.value
  
  if (p_value < 0.001) {
    p_label <- "p < 0.001"
  } else {
    p_label <- paste0("p = ", round(p_value, 3))
  }
  
  # Create the plot with the corresponding unit
  p <- ggplot(data_g2_g3, aes_string(x = "Grupo", y = variables[i], fill = "Grupo")) +
  geom_violin(trim = FALSE, width = 1.2, color = "black", alpha = 0.6) +
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.7, color = "black") +
  stat_summary(fun = "median", geom = "point", size = 2, color = "black") + 
  labs(x = NULL, y = units[i], title = variables[i]) + 
  annotate("text", x = 1.5, y = Inf, label = p_label, vjust = 2, size = 8, fontface = "plain") +
  theme_minimal(base_size = 14) + 
  scale_fill_manual(values = c("#6AB1F4", "#F26553"), guide = "none") + 
  scale_x_discrete(labels = levels(data_g2_g3$Grupo)) + 
  theme(
    legend.position = "none", 
    axis.title = element_text(size = 20), 
    axis.text.x = element_text(size = 20), 
    axis.ticks.x = element_blank(),
    plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),
    plot.margin = margin(5, 5, 10, 5) 
  )
  
  plot_list[[variables[i]]] <- p
}

combined_plot <- ggarrange(plotlist = plot_list, ncol = 5, nrow = 2)

print(combined_plot)

# Save the combined image
ggsave("Violin_Plots_XGB_g2g3.png", combined_plot, width = 20, height = 10, bg = "white")
```


